First results of developing a Swedish parser based on a wide coverage grammar

### 2011-10-10 doc/paper/Paper.tex
things

\title{First results of developing a ruled based Swedish parser} %A Wide-Coverage Swedish Grammar} % ska vi vara anonyma?
\author{Malin Ahlberg
% vad ska stå här eg?
\institute{Center of Language Technology, Gothenburg University, Sweden}
%\email{ahlberg.malin@gmail.com}
}
\def\titlerunning{First results of developing a ruled based Swedish parser}
\def\authorrunning{M. Ahlberg}
\begin{document}
\maketitle

\begin{abstract}
This paper describes an on-going work of a rule-based, open source
parser for Swedish, based on a wide-coverage grammar implemented in
Grammatical Framework (GF).
In addition to GF, we use two other main resources; the Swedish treebank Talbanken
and the electronic lexicon SALDO.
The grammar will be highly reusable as GF provides libraries for compiling
grammars to a number of programming languages.
% allow easy interaction with other programs and grammars
%

%>> something very commercial, interesting and outstanding here please!

@@

%R: -- the second paragraph - OK but please extend.
We would like to implement a wide coverage grammar and parser for Swedish.
We thereby investigate how the grammar formalism Grammatical Framework,
further described in section \ref{sec:gf}, can be used for open domain parsing.
By starting from an already existing Swedish grammar written in GF, we get 
an fundamental description of the language. The framework also provides
tools such as parsing, generation and
a well tested interpretation of the parse trees. There are also tools for using
the grammar in a number of programming languages. 
%>> , tools for visualizing ...

@@

%We aim to implement % The goal is to be able to 
%a parser for open domain language and 
%This paper describes our work and results 
%so far as well as the directions of future work.

\section{Background}
\subsection{Grammatical Framework}
\label{sec:gf}

@@

Grammatical Framework\cite{gf} (GF) is a grammar formalism based on functional
programming, designed for multilingual grammar applications.

The key idea
is to divide the grammar into abstract and concrete.
The abstract grammar gives a logical representation of the semantics of the
domain, modeled as abstract trees. Complications occurring in
natural languages, such as agreement, case and word order, are abstracted away. 
%R: what complications do you abstract away ?
%categories and trees are not interesting as such, but their usage
%is - why do you need them in the abstract syntax actually ?
That is, the abstract grammar contains no actual linguistic information, but
information about what categories are used and how to combine them into trees.

@@

% obra mening
The grammar acts as an independent module and may be used in different
projects.
%R: more about the GF library - if you have space
Reusability is further supported by the GF resource grammars\cite{resource},
a library provided with the GF package, 
where information about the morphological and
syntactical rules for more than 20 different languages are implemented.
As a result, an end user of the framework does not need to describe how to 
form standard sentences, phrases or decline words, but can focus on the
particularities for the given domain and choose what kind of 
utterances that should be allowed.

%so far bara used för vissa saker, 
% about why we need more, extra module, separate swedish from scandinavian
% get inspired by Shafqat.
GF has so far been used in a number of projects, MOLTO\cite{molto} and
WebAlt\cite{webalt} to mention a few. 
All those are special domain applications, dealing with controlled natural
language.
This project experiments on other usages of a GF grammar.

@@

\subsection{Talbanken}
%Wherefrom, how big, who, when, version, used for maltparser
Talbanken\cite{talbanken} is
a Swedish treebank put together in the 1970s at Lund University.
In 2005 it was modernized by Nivre, Nilsson and Hall\cite{talbanken05} and
enriched with annotation for a
full phrase structure analysis. 
Talbanken contains both written and spoken Swedish. For this project only
the prose material, consisting of 6316 sentences will be used.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\
We are using parts of Talbanken for testing, development and evaluation.
%and the parsetree generated by our grammar will eventually be 
%compared to those from the treebank.

@@

\section{Work in progress}
\subsection{A tool for lexical acquisition}
%R: lexicon extraction part - more about everything, comparison with
%Extract and Functional Morphology, evaluation of the lexicon builder -
%"supervised lexicon extraction"(common term for this - google it if
%you want to find related work)

To enlarge the lexicon, 
a tool for automatically acquisition has been created. It has 
been tested on verbs with good results. It makes use of
the \emph{smart paradigm} given in the Swedish resource grammar.
The smart paradigm accepts verbs in present tense indicative form and decides
 which paradigm it most likely belong to. If needed, it also accepts more
verb forms showing the correct inflection.
%which Making use of the

By combining this method with the information from the tags in Talbanken,
%and the paradigms in the Swedish resource grammar, 
the tool interactively generates GF lexicons. 
Given a list of words, it iteratively
tries to figure out how to conjugate each of them. If several forms of a word is 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection table. 
If this table contains all other forms from the input list,
the program will ask the  user to
validate the claimed conjunction. The user may now either
allow the word to be added to the lexicon, remove it or demand the program
to make another guess.
Although using simple techniques, the tool 
manage to correctly guess 70-75\% of the given lemmas when tested on
verbs. \\
For further extractions, there are also earlier developed tools such as
Extract\cite{extract} for supervised lexicon extraction and
FM\cite{fm} for programming lexical resources. Extract uses a similar 
method but does not have access to the GF smart paradigm.

@@

For multilingual GF applications, translation should always be possible, and
therefore the abstract may only contain constructions that
are common to all implemented languages.
%>> resources for controlled language. does not aim for covering whole language
%>> googla efter bra citat
This means that everything added to the resources
must be expressable in all other languages.
Therefore, the resources have to be general and non language specific.
Language specific constructs such as stylistic changes, idioms, informal
expressions etc may be given in a special module.
This module, together with the Swedish resource grammar has been the starting
point of this project.

Since many of the languages in the GF library reassemble each other grammatically,
they can share much of their grammar implementations. This is usually done by using a
\verb|Functor|, which lets a number of languages share parts of their implementation. 
In addition to simply avoiding code duplication, this technique aids the code maintenance.
For Swedish, about 85\% of the resource code is shared with the other Scandinavian
languages.
%due to the resemblances between the languages, they share
%>> using a functor for sharing code, for when having the same structure but
%>> but only different words. 
For a bigger and still well-defined Swedish grammar however, the implementation of
the languages needs to be more separated. % in order to avoid ad-hock solutions.
%<<in order to avoid ad-hoc solutions and 
Some constructions that are standard in
Norwegian and Danish may be very rare in Swedish and while extending the
grammar there may occur situations where a deeper restructuring of the
implementation is needed. \\
A number of grammatical constructions have been added the the Swedish grammar.
%>> this is language specific, therefore not in common abstract
Those include constructs that are generally not present in other languages,
such as the reflexive pronoun \emph{sitt}: \\
\emph{Han såg \textbf{sitt} hus} $\;$ (\emph{He saw his (own) house}) $\;$
as opposed to \\
\emph{Han såg \textbf{hans} hus} $\;$ (\emph{He saw his (an other person's) house}) \\
%>> an stylistic change that has been added, not needed for expressable new things (typ) 
%>> not critical, does not have to be in resource grammar
%>> also not expressable in all languages
The possibility to put a part of a sentence in focus has also been added:\\
\emph{Glad var han inte.} $\;$ (\emph{Happy was he not}). \\
This sort of rephrasing which has little effect on the logical representation
is commonly not given by the resource grammar.

In order to keep the grammar as clean and to keep it from being too allowing,
much care is taken to develop this as neatly as possible. 
%<<Some ellipses are added, while there is work going on to add trickier ones (eller inte 
%<<ellipser just nu dårå, men vi skriver något fint här med, ja-jaaaa!).

@@

After having developed the grammar and lexicon, we plan to 
make the parser robust by using techniques such as chunk parsing, named
entity recognition, methods for handling unknown words and
grammatical constructions, idioms, ellipses etc.

@@

\section{Acknowledgments}
Special thanks to Aarne Ranta, Ramona Enache, Elisabet Engdahl, Krasimir
Angelov and Lars Borin for their help and support.

>>>>

\title{First results of developing a Swedish parser based on a wide coverage grammar} %A Wide-Coverage Swedish Grammar} % ska vi vara anonyma?
\author{Malin Ahlberg
% vad ska stå här eg?
\institute{Center of Language Technology, Gothenburg University, Sweden}
%\email{ahlberg.malin@gmail.com}
}
\def\titlerunning{First results of developing a grammar based Swedish parser}
\def\authorrunning{M. Ahlberg}
\begin{document}
\maketitle

\begin{abstract}
This paper describes an on-going work of a rule-based, open source
parser for Swedish. The central component will be a wide-coverage grammar
implemented in the grammar formalism Grammatical Framework (GF). The
resulting parser will be the first deep Swedish parser based on a grammar and
will give a full syntactic analysis of the input sentences.
In addition to GF, we use two other main resources; the Swedish
treebank Talbanken and the electronic lexicon SALDO.
% ha med detta?
The grammar will be highly reusable, chosen 
parts of the it can be used for applications dealing with controlled 
or free natural language. GF also provides libraries for compiling
grammars to a number of programming languages.
% allow easy interaction with other programs and grammars
% 
%>>
%>> something very commercial, interesting and outstanding here please!

@@

Swedish is a North Germanic language spoken by approximately 10 million people.
It is a SVO language, and although the word order is relatively strict,
inverted order is commonly used
to indicate questions or to emphazise different parts of the sentence. \\
Language technology for Swedish is an area of much interesting research.
%on  are currently being conducted. \\
%R: -- the second paragraph - OK but please extend.
Our goal is to implement a wide coverage grammar and parser for Swedish.
We thereby investigate how the grammar formalism Grammatical Framework
%further described in section \ref{sec:gf}, 
can be used for open domain parsing.
By starting from an already existing Swedish grammar written in GF, we get 
an fundamental description of the language. The framework also provides
tools such as parsing, generation and
a well tested interpretation of the parse trees. Furthermore, there are tools
for using the grammar in a number of programming languages. 
%>> , tools for visualizing ...

@@

This paper will briefly introduce GF, Talbanken and SALDO in section \ref{sec:background}
and describe our current work in section \ref{sec:progress}.
%We aim to implement % The goal is to be able to 
%a parser for open domain language and 
%This paper describes our work and results 
%so far as well as the directions of future work.

\section{Background}
\label{sec:background}
\subsection{Grammatical Framework}
\label{sec:gf}

@@

Grammatical Framework\cite{ranta-2011} (GF) is a grammar formalism based on functional
programming, designed for multilingual grammar applications.

The key idea
is to divide the grammar into abstract and concrete parts. %% parts?
The abstract grammar gives a logical representation of the semantics,
%of the domain, 
modeled as abstract trees. Complications occurring in
natural languages, such as agreement, case and word order, are abstracted away. 
%R: what complications do you abstract away ?
%categories and trees are not interesting as such, but their usage
%is - why do you need them in the abstract syntax actually ?
That is, the abstract grammar contains no actual linguistic information, but
information about what categories are used and how to combine them into trees.

@@

The grammar acts as an independent module and may be used in different
projects.
%R: more about the GF library - if you have space
Reusability is further supported by the separation between resource grammars
and appliciation grammars. The resource grammars is 
a library provided with the GF package, 
where information about the morphological and
syntactical rules for more than 20 different languages are implemented.
Hence a writer of a application grammar can start her work at a higher
level and does not need to describe how to 
form standard sentences, phrases or decline words.
%, but can focus on the
%particularities for the given domain and choose what kind of 
%utterances that should be allowed.

%so far bara used för vissa saker, 
% about why we need more, extra module, separate swedish from scandinavian
% get inspired by Shafqat.
GF has so far been used in a number of projects, MOLTO\cite{molto}, TALK\cite{talk}
and WebAlt\cite{webalt} to mention a few. 
All those are special domain applications, dealing with controlled natural
language.
This project experiments on other usage of a GF grammar - to use it for open
domain language.

@@

\subsection{Talbanken}
%Wherefrom, how big, who, when, version, used for maltparser
Talbanken\cite{talbanken} is
a Swedish treebank put together in the 1970s at Lund University.
In 2005 it was modernized by Nivre, Nilsson and Hall\cite{talbanken05} and
enriched with annotation for a full phrase structure analysis. 
We are using the treebank for development and evaluation.

Although Talbanken contains both written and spoken Swedish,
%For this project 
only the prose material, consisting of 6316 sentences will be used in this project.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\
%and the parsetree generated by our grammar will eventually be 
%compared to those from the treebank.

@@

\section{Work in progress}
\label{sec:progress}
\subsection{A tool for lexical acquisition}
%R: lexicon extraction part - more about everything, comparison with
%Extract and Functional Morphology, evaluation of the lexicon builder -
%"supervised lexicon extraction"(common term for this - google it if
%you want to find related work)

To enlarge the lexicon, 
a tool for automatically acquisition has been created. It has 
been tested on verbs with good results. It makes use of
the \emph{smart paradigm} given in the Swedish resource grammar.
The smart paradigm acts as functions that given one form of a word can
infer which paradigm it most likely belong to.
The smart paradigm for verbs accepts words in present tense indicative form.
If needed, it also accepts more verb forms showing the correct inflection.
%which Making use of the

By combining this method with the information from the tags in Talbanken,
%and the paradigms in the Swedish resource grammar, 
the tool interactively generates GF lexicons. 
Given a list of words, it iteratively
tries to figure out how to conjugate each of them. If several forms of a word is 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection table. 
If this table contains all other forms from the input list,
the program will ask the  user to
validate the claimed conjunction. The user may now either
allow the word to be added to the lexicon, remove it or demand the program
to make another guess.
Although using simple techniques, the tool 
manage to correctly guess 70-75\% of the given lemmas when tested on
verbs. \\
For further extractions, there are also earlier developed tools such as
Extract\cite{extract} for supervised lexicon extraction and
FM\cite{fm} for programming lexical resources.
%Extract uses a similar 
%method but defines the paradigms itself.
%rules are given in another format.

@@

For multilingual GF applications, translation should always be possible, and
therefore the resource abstract may only contain constructions that
are common to all implemented languages.
%>> resources for controlled language. does not aim for covering whole language
%>> googla efter bra citat
%This means that everything added to the resources must be expressable in all other languages.
Therefore, the resources have to be general and non language specific.
Language specific constructs such as stylistic changes, idioms, informal
expressions etc may be given in a special module.
This module, together with the Swedish resource grammar has been the starting
point of this project.

Since many of the languages in the GF library reassemble each other grammatically,
they can share much of their grammar implementations. This is usually done by using a
\verb|Functor|, which lets a number of languages share parts of their implementation. 
In addition to simply avoiding code duplication, this technique aids the code maintenance.
For Swedish, about 85\% of the resource code is shared with the other Scandinavian
languages. % >> When trying to achieve a lightweight, usable code? Not
%due to the resemblances between the languages, they share
%>> using a functor for sharing code, for when having the same structure but
%>> but only different words. 
However, if we aim for a deeper and more covering analysis of Swedish,
%For a bigger and still well-defined Swedish grammar however,
the implementation of the languages needs to be more separated. 
% in order to avoid ad-hock solutions.
%<<in order to avoid ad-hoc solutions and 
%Some constructions that are standard in
%Norwegian and Danish may be very rare in Swedish and while extending the
%grammar there may occur situations where a deeper restructuring of the
%implementation is needed. \\
A number of grammatical constructions have been added the the Swedish grammar.
%>> this is language specific, therefore not in common abstract
Those include constructs that are generally not present in other languages,
such as the reflexive pronoun \emph{sitt}: \\
\emph{Han såg \textbf{sitt} hus} $\; \; \; \;$ (\emph{He saw his (own) house}) $\;$
as opposed to \\
\emph{Han såg \textbf{hans} hus} $\;$ (\emph{He saw his (an other person's) house}) \\
%>> an stylistic change that has been added, not needed for expressable new things (typ) 
%>> not critical, does not have to be in resource grammar
%>> also not expressable in all languages
The possibility to put a part of a sentence in focus has also been added:\\
\emph{Glad var han inte.} $\;$ (\emph{Happy was he not}). \\
This sort of rephrasing which has little effect on the logical representation
is commonly not given by the resource grammar.

In order to keep the grammar as clean and to keep it from allowing syntactical errors,
much care is taken to develop this as neatly as possible. 
%<<Some ellipses are added, while there is work going on to add trickier ones (eller inte 
%<<ellipser just nu dårå, men vi skriver något fint här med, ja-jaaaa!).

\section{Related work}
There are other parser for Swedish parser such as the the statistical MaltParser\cite{malt}
which was trained on Talbanken. 
CassSwe (Kokkinakis and Johansson-Kokkinakis, 1999) is a  based on finite state cascades,
whereas shallow parser GTA (Knutsson, Bigert and Kann, 2003) relies on rules of 
a context free grammar. Both GTA and CassSwe operates on POS tagged text.
%deep Swedish Core Language Parser,
%shallow Swedish Constraint Grammar.

@@

After having developed the grammar and lexicon, we plan to 
make the parser robust by using techniques such as chunk parsing, named
entity recognition, methods for handling unknown 
grammatical constructions such as idioms and ellipses. %etc.

@@

\section{Acknowledgments}
The work has been funded by Center of Language Technology.
I would also like to give special thanks to Aarne Ranta, Ramona Enache,
Elisabet Engdahl, Krasimir Angelov and Lars Borin for their help and support.
